### 2014-02-17 thesis-proposal/acc.tex
Updated proposal

In this work, we discuss active code completion in the context of object construction in Java because type-aware editors for Java are better than developed than those for other languages, because we wish to do user studies, and because Java already provides a way to associate metadata with classes. The techniques in this section apply equally well to type-aware editors for any language with a similar mechanism.

>>>>

In this work, we discuss active code completion in the context of object construction in Java because type-aware editors for Java are better developed than those for other languages, because we wish to do empirical studies, and because Java already provides a way to associate metadata with classes. The techniques in this section apply equally well to type-aware editors for any language with a similar mechanism.

### 2014-02-17 thesis-proposal/background.tex
Updated proposal

% This stuff isn't actually hooking into the compiler...
The first concrete realizations of active libraries in statically typed settings, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation. A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{veldhuizen2000blitz++}. Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about (see discussion in \cite{Robison:2001:IEC:376656.376751}). %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.

>>>>

% This stuff isn't actually hooking into the compiler...
The first concrete realizations of active libraries in statically typed settings, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation. A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{veldhuizen2000blitz++}. Although this and several other interesting optimizations are possible by this technique, its expressiveness is fundamentally limited because template expansion allows for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about (see discussion in \cite{Robison:2001:IEC:376656.376751}). %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.

### 2014-02-17 thesis-proposal/parsing.tex
Updated proposal

% !TEX root = omar-thesis-proposal.tex
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\rmfamily, 
stringstyle=\ttfamily,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,valAST},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
moredelim=**[is][\color{blue}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID}
}
\lstloadlanguages{Java,VBScript,XML,HTML}
\let\li\lstinline

@@

When a specialized notation like this is not available, but the equivalent general-purpose notation is too cognitively demanding for comfort, developers typically turn to run-time mechanisms to make constructing data structures more convenient. Among the most common strategies in these situations is to simply use a string representation that is parsed at run-time. Developers across language paradigms frequently write examples like those above as:
\begin{lstlisting}
let webpage : HTML = parse_html($\texttt{"}$SSTR<html><body><h1>Results for ESTR$\texttt{"}$ + keyword + $\texttt{"}$SSTR</h1>
  <ul id=\$\texttt{\color{cyan}"}$results\$\texttt{\color{cyan}"}$>$\texttt{"}$ESTR + to_string(to_list_items(query(db, parse_sql(
  	$\texttt{"}$SSTRSELECT title, snippet FROM products WHERE '$\texttt{" + keyword + "}$' in title$\texttt{"}$ESTR)))) + 
  $\texttt{"}$SSTR</ul></body></html>$\texttt{")}$
\end{lstlisting}

@@

In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker. The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against. The TSL is responsible for rewriting this term to ultimately use only general-purpose notation. This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly. It also avoids semantic conflicts -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, dictionaries and other data structures, like JSON literals. This also frees these common notations from being tied to the variant of a  data structure built into the standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
%\begin{lstlisting}
%let empty_set : Set = { }
%let empty_dict : Dict = { }
%let empty_json : JSON = { }
%\end{lstlisting}
\subsection{Wyvern}
We develop our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}. To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects or mutable state) and it does not enforce a uniform access principle for objects (fields can be accessed directly). Objects are thus essentially labeled product types with simple methods (functions that require a self-reference). We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML. One can refer to the version of the language described in this thesis as \emph{TSL Wyvern} when the variant is not clear.

\subsection{Example: Web Search}
We begin in Fig. \ref{f-example} with an example showing several different TSLs being used to define a fragment of a web application showing search results from a database. We will go through this example below to develop intuitions about TSLs in Wyvern. Note that for clarity of presentation, we color each character  according to the TSL it is governed by.
\begin{figure}[t]
\begin{lstlisting}
let imageBase : URL = <SURLimages.example.comEURL>
let bgImage : URL = <SURL%EURLimageBaseSURL%/background.pngEURL>
new : SearchServer
  def resultsFor(searchQuery : String, page : Nat) : Unit =
    serve(~) (* serve : HTML -> Unit *)
SHTML      :html
        :head
          :title Search Results
          :style EHTML~
SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
            #search { background-color: %ECSS`SCOLOR#aabbccECOLOR`.darken(SPCT20pctEPCT)SCSS% }
ECSSSHTML        :body
          :h1 Results for {EHTMLsearchQuerySHTML}
          :div[id="search"]
            Search again: {EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML}
          {EHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
            fmt_results(db, ~, SNAT10ENDNAT, page)
              SSQLSELECT * FROM products WHERE {ESQLsearchQuerySSQL} in titleESQL
          SHTML}
\end{lstlisting}
\vspace{-8px}
\caption{Wyvern Example with Multiple TSLs}
\label{f-example}
\vspace{-10px}
\end{figure}
\subsection{Inline Literals}
\begin{figure}[t]
\begin{lstlisting}
<SURLliteral body here, <inner angle brackets> must be balancedEURL>
{SURLliteral body here, {inner braces} must be balancedEURL}
[SURLliteral body here, [inner brackets] must be balancedEURL]
`SURLliteral body here, ``inner backticks`` must be doubledEURL`
$\texttt{'}$SURLliteral body here, ''inner single quotes'' must be doubledEURL$\texttt{'}$
$\texttt{"}$SURLliteral body here, ""inner double quotes"" must be doubledEURL$\texttt{"}$
SURL12xyzEURL (* no delimiters necessary for number literals; suffix optional *)
\end{lstlisting}
\vspace{-8px}
\caption{Inline Generic Literal Forms}
\vspace{-10px}
\label{f-delims}
\end{figure}
Our first TSL appears on the right-hand side of the variable binding on line 1. The variable \li{imageBase} is annotated with its type, \li{URL} (not shown). This is an {object type} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on. We could have created a value of type \li{URL} using general-purpose notation\todo{fix comment formatting}\todo{new color}:
\begin{lstlisting}
let imageBase : URL = new
  val protocol : URLString = $\text{"}$SSTRhttpESTR$\texttt{"}$
  val subdomain : URLString = $\texttt{"}$SSTRimagesESTR$\texttt{"}$
  (* ... *)
\end{lstlisting}
%  val domain : URLString = $\texttt{"}$SSTRexampleESTR$\texttt{"}$
This is tedious. Instead, because the \li{URL} type has a TSL associated with it, we can instead introduce precisely this value using conventional notation for URLs by placing it in the \emph{body} of a \emph{generic literal}, \li{<SURLimages.example.comEURL>}. Any other delimited form in Fig. \ref{f-delims} could equivalently be used if the constraints shown are obeyed. The type annotation on \li{imageBase} implies that this literal's \emph{expected type} is \li{URL}, so the {body} of the literal (the characters between the angle brackets, in blue) will be governed by the \li{URL} TSL during the typechecking phase. This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL}.

@@

Each production is followed, in an indented block, by a Wyvern function that generates a value given the values recursively generated by each of the $n$ non-terminals in contains, ordered left-to-right. For the starting non-terminal, always called \li{start}, this function must return a value of type \li{Exp}. User-defined non-terminals might have a different type associated with them (not shown). Here, we show how to generate an AST using general-purpose notation for \li{Exp} (lines 13-15) as well as a more natural \emph{quasiquote} style (lines 11 and 18). Quasiquotes are expressions that are not evaluated, but rather reified into syntax trees. We observe that quasiquotes too fall into the pattern of ``specialized notation associated with a type'' -- quasiquotes for expressions, types and identifiers are simply TSLs associated with \li{Exp}, \li{Type} and \li{ID} (Fig. \ref{f-builtins}). They support the full Wyvern concrete syntax as well as an additional delimited form, written with \li{%}s, that supports ``unquoting'': interpolating another AST into the one being generated. Again, interpolation is type safe and structured, rather than based on string interpolation.

@@

\begin{figure}
\begin{lstlisting}[escapechar=$]
casetype HTML 
    Empty of Unit
  | Text of String
  | Seq of HTML * HTML 
  | BodyElement of Attributes * HTML
  | StyleElement  of Attributes * CSS
  | ...
  metadata = new : HasTSL
    val parser : Parser = ~
SGRM      start -> ':body'= start>
        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
SGRM      start -> ':style'= EXP>
        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
SGRM      start -> '{'= EXP '}'=EGRM
        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
\end{lstlisting}
\vspace{-8px}
\caption{A Wyvern case type with an associated TSL.}
\label{f-htmltype}
\end{figure}
\begin{figure}[t]
\begin{subfigure}[t]{.53\textwidth}
\begin{lstlisting}
objtype HasTSL
  val parser : Parser
objtype Parser                          
  def parse(ts : TokenStream) : (Exp * 
    TokenStream)
  metadata = new : HasTSL
    val parser : Parser = (* parser generator *)      
casetype Type
    Var of ID
  | Arrow of Type * Type
  | Prod of Type * Type
  metadata = new : HasTSL
    val parser : Parser = (* type quasiquotes *)
\end{lstlisting}
\end{subfigure}
\begin{subfigure}[t]{.55\textwidth}
\begin{lstlisting}[linewidth=.5\textwidth]
casetype Exp 
    Var of ID
  | Lam of ID * Type * Exp
  | Ap of Exp * Exp
  | ProdIntro of Exp * Exp
  | CaseIntro of Type * ID * Exp
  ... 
  | FromTS of Exp * Exp
  | Literal of TokenStream
  | Error of ErrorMessage
  metadata = new : HasTSL
    val parser : Parser = (* quasiquotes *)
\end{lstlisting}
\end{subfigure}
\caption{Some of the types included in the Wyvern prelude.}
\vspace{-10px}
\label{f-builtins}
%\label{fig:typeParser}
\end{figure}

\subsection{Formalization}
A formal and more detailed description can be found in our paper draft\footnote{\url{https://github.com/wyvernlang/docs/tree/master/ecoop14}}. In particular:
\begin{enumerate}
\item We provide a complete layout-sensitive concrete syntax. We show how it can be written without the need for a context-sensitive lexer or parser using Adams' formalism and provide a full specification for the layout-delimited literal form introduced by a forward reference, \li{~}, as well as other forms of forward-referenced blocks (for the forms \li{new} and \li{case(e)}, in particular).
%\item We detail the general mechanism for associating metadata with a type. A TSL is then implemented by associating a parser (of type \li{Parser}) with a type. The parser is responsible for rewriting tokenstreams (of type \li{Tokenstream}) into Wyvern ASTs (of type \li{Exp}). These types are defined in the standard library.
%\item This lower-level mechanism is general, but writing a hand-written parser and manipulating syntax trees manually is cognitively demanding. We observe that \emph{grammars} and \emph{quasiquotes} can both be seen as TSLs for parsers and ASTs respectively and discuss how to implement them as such.
\item We formalize the static semantics, including the literal parsing logic, of TSL Wyvern by combining a bidirectional type system (in the style of Lovas and Pfenning \cite{lovasPfenning}) with an elaboration semantics (in the style of Harper and Stone \cite{harperStone}). By distinguishing locations where an expression synthesizes a type from locations where an expression is being analyzed against a known type, we can precisely state where generic literals can and cannot appear and how parsing is delegated to a TSL.
\item A na\"ive rewriting strategy would be \emph{unhygienic} -- it could allow for the inadvertent capture of local variables. We show a novel mechanism that ensures hygiene by requiring that the generated AST is closed except for subtrees derived from portions of the user's tokenstream that are interpreted as nested Wyvern expressions. We also show how to explicitly refer to local values available in the parser definition (e.g. helper functions) in a safe way. We formalize this hygiene mechanism by bifurcating the context in our static semantics.
\item We provide several examples of TSLs throughout the paper, but to examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova\todo{and Josh?}).
\end{enumerate}

\subsection{Remaining Tasks and Timeline}
The paper we have submitted for ECOOP 2014 has received quite positive reviews, so we anticipate that it will be accepted (notifications are sent on Mar. 7th). Other than final revisions on the paper text, due on May 12th, the following tasks remain to be completed. We will largely do these tasks after the OOPSLA deadline in late March.
\begin{enumerate}
%\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints. This is not strictly necessary for preventing ambiguities, and it is quite convenient to not require this (as in Python, for example). We will add these simple rules to the grammar.
\item We must write down the full formal semantics (including rules that we omitted for concision in the paper draft) in a technical report, as well as provide more complete proofs of the metatheory. 
\item Our current static semantics does not support mutually recursive type declarations, but this is necessary for our prelude to typecheck, so we will add support for this.
\item We must further consider aspects of hygiene. In particular, we do not yet have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}. It may be possible to prevent shadowing by making it impossible for variables to be introduced into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL we have shown).
%\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
%\item The current implementation of Wyvern does not include all of these mechanisms as described. Although the implementation of Wyvern is not a project I am responsible for, I plan on working with the student who is doing this to implement as much of what we have described as possible.\todo{should I drop this from the proposal? I don't want to promise that everything will be implemented because that is largely contingent on Benjamin's effort, not mine.}
\end{enumerate}
%
%\subsection{TODO}
%- Figure out separate compilation (e.g. of CSS)

>>>>

% !TEX root = omar-thesis-proposal.tex
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily, 
stringstyle=\ttfamily,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,valAST},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
moredelim=**[is][\color{blue}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS}
}
\lstloadlanguages{Java,VBScript,XML,HTML}
\let\li\lstinline

@@

When a specialized notation like this is not available, but the equivalent general-purpose notation is too cognitively demanding for comfort, developers typically turn to run-time mechanisms to make constructing data structures more convenient. Among the most common strategies in these situations is to simply use a string representation that is parsed at run-time. Developers across language paradigms frequently write examples like the above as:
\begin{lstlisting}
let webpage : HTML = parse_html($\texttt{"}$SSTR<html><body><h1>Results for ESTR$\texttt{"}$ + keyword + $\texttt{"}$SSTR</h1>
  <ul id=\$\texttt{\color{cyan}"}$results\$\texttt{\color{cyan}"}$>$\texttt{"}$ESTR + to_string(to_list_items(query(db, parse_sql(
  	$\texttt{"}$SSTRSELECT title, snippet FROM products WHERE '$\texttt{" + keyword + "}$' in title$\texttt{"}$ESTR)))) + 
  $\texttt{"}$SSTR</ul></body></html>$\texttt{")}$
\end{lstlisting}

@@

In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker. The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against. The TSL is responsible for rewriting this term to ultimately use only general-purpose notation. This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly. It also avoids semantic conflicts -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, dictionaries and other data structures, like JSON literals. This also frees these common notations from being tied to the variant of a  data structure built into the standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
%\begin{lstlisting}
%let empty_set : Set = { }
%let empty_dict : Dict = { }
%let empty_json : JSON = { }
%\end{lstlisting}
\subsection{Wyvern}
We develop our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}. To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects or mutable state) and it does not enforce a uniform access principle for objects (fields can be accessed directly). Objects are thus essentially recursive labeled product types with simple methods (functions that require a self-reference). We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML. One can refer to the version of the language described in this thesis as \emph{TSL Wyvern} when the variant is not clear.

\subsection{Example: Web Search}
We begin in Fig. \ref{f-example} with an example showing several different TSLs being used to define a fragment of a web application showing search results from a database. Note that for clarity of presentation, we color each character  according to the TSL it is governed by.
\begin{figure}[t]
\begin{lstlisting}
let imageBase : URL = <SURLimages.example.comEURL>
let bgImage : URL = <SURL%EURLimageBaseSURL%/background.pngEURL>
new : SearchServer
  def resultsFor(searchQuery : String, page : Nat) : Unit =
    serve(~) (* serve : HTML -> Unit *)
SHTML      :html
        :head
          :title Search Results
          :style EHTML~
SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
            #search { background-color: %ECSS`SCOLOR#aabbccECOLOR`.darken(SPCT20pctEPCT)SCSS% }
ECSSSHTML        :body
          :h1 Results for {EHTMLsearchQuerySHTML}
          :div[id="search"]
            Search again: {EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML}
          {EHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
            fmt_results(db, ~, SNAT10ENDNAT, page)
              SSQLSELECT * FROM products WHERE {ESQLsearchQuerySSQL} in titleESQL
          SHTML}
\end{lstlisting}
\vspace{-8px}
\caption{Wyvern Example with Multiple TSLs}
\label{f-example}
\vspace{-10px}
\end{figure}
\subsection{Inline Literals}
\begin{figure}[t]
\begin{lstlisting}
<SURLliteral body here, <inner angle brackets> must be balancedEURL>
{SURLliteral body here, {inner braces} must be balancedEURL}
[SURLliteral body here, [inner brackets] must be balancedEURL]
`SURLliteral body here, ``inner backticks`` must be doubledEURL`
$\texttt{'}$SURLliteral body here, ''inner single quotes'' must be doubledEURL$\texttt{'}$
$\texttt{"}$SURLliteral body here, ""inner double quotes"" must be doubledEURL$\texttt{"}$
SURL12xyzEURL (* no delimiters necessary for number literals; suffix optional *)
\end{lstlisting}
\vspace{-8px}
\caption{Inline Generic Literal Forms}
\vspace{-10px}
\label{f-delims}
\end{figure}
Our first TSL appears on the right-hand side of the variable binding on line 1. The variable \li{imageBase} is annotated with its type, \li{URL} (not shown). This is an {object type} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on. We could have created a value of type \li{URL} using general-purpose notation:
\begin{lstlisting}
let imageBase : URL = new
  val protocol : URLString = $\text{"}$SUShttpEUS$\texttt{"}$
  val subdomain : URLString = $\texttt{"}$SUSimagesEUS$\texttt{"}$
  (* ... *)
\end{lstlisting}
%  val domain : URLString = $\texttt{"}$SSTRexampleESTR$\texttt{"}$
This is tedious. Instead, because the \li{URL} type has a TSL associated with it, we can instead introduce precisely this value using conventional notation for URLs by placing it in the \emph{body} of a \emph{generic literal}, \li{<SURLimages.example.comEURL>}. Any other delimited form in Fig. \ref{f-delims} could equivalently be used if the constraints shown are obeyed. The type annotation on \li{imageBase} implies that this literal's \emph{expected type} is \li{URL}, so the {body} of the literal (the characters between the angle brackets, in blue) will be governed by the \li{URL} TSL during the typechecking phase. This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL}.

@@

Each production is followed, in an indented block, by a Wyvern function that generates a value given the values recursively generated by each of the $n$ non-terminals it contains, ordered left-to-right. For the starting non-terminal, always called \li{start}, this function must return a value of type \li{Exp}. User-defined non-terminals might have a different type associated with them (not shown). Here, we show how to generate an AST using general-purpose notation for \li{Exp} (lines 13-15) as well as a more natural \emph{quasiquote} style (lines 11 and 18). Quasiquotes are expressions that are not evaluated, but rather reified into syntax trees. We observe that quasiquotes too fall into the pattern of ``specialized notation associated with a type'' -- quasiquotes for expressions, types and identifiers are simply TSLs associated with \li{Exp}, \li{Type} and \li{ID} (Fig. \ref{f-builtins}). They support the full Wyvern concrete syntax as well as an additional delimited form, written with \li{%}s, that supports ``unquoting'': interpolating another AST into the one being generated. Again, interpolation is type safe and structured, rather than based on string interpolation.

@@

\begin{figure}
\begin{lstlisting}[escapechar=$]
casetype HTML 
    Empty of Unit
  | Text of String
  | Seq of HTML * HTML 
  | BodyElement of Attributes * HTML
  | StyleElement  of Attributes * CSS
  | ...
  metadata = new : HasTSL
    val parser : Parser = ~
SGRM      start -> ':body'= start>
        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
SGRM      start -> ':style'= EXP>
        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
SGRM      start -> '{'= EXP '}'=EGRM
        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
\end{lstlisting}
\vspace{-8px}
\caption{A Wyvern case type with an associated TSL.}
\label{f-htmltype}
\end{figure}
\begin{figure}[t]
\begin{subfigure}[t]{.53\textwidth}
\begin{lstlisting}
objtype HasTSL
  val parser : Parser
objtype Parser                          
  def parse(ts : TokenStream) : (Exp * 
    TokenStream)
  metadata = new : HasTSL
    val parser : Parser = (* parser generator *)      
casetype Type
    Var of ID
  | Arrow of Type * Type
  | Prod of Type * Type
  metadata = new : HasTSL
    val parser : Parser = (* type quasiquotes *)
\end{lstlisting}
\end{subfigure}
\begin{subfigure}[t]{.55\textwidth}
\begin{lstlisting}[linewidth=.5\textwidth]
casetype Exp 
    Var of ID
  | Lam of ID * Type * Exp
  | Ap of Exp * Exp
  | ProdIntro of Exp * Exp
  | CaseIntro of Type * ID * Exp
  ... 
  | FromTS of Tokenstream
  | Literal of TokenStream
  | Error of ErrorMessage
  metadata = new : HasTSL
    val parser : Parser = (* quasiquotes *)
\end{lstlisting}
\end{subfigure}
\caption{Some of the types included in the Wyvern prelude.}
\vspace{-10px}
\label{f-builtins}
%\label{fig:typeParser}
\end{figure}

\subsection{Formalization}
A formal and more detailed description can be found in our paper draft.\footnote{\url{https://github.com/wyvernlang/docs/blob/master/ecoop14/ecoop14.pdf?raw=true}} In particular:
\begin{enumerate}
\item We provide a complete layout-sensitive concrete syntax. We show how it can be written without the need for a context-sensitive lexer or parser using Adams' formalism and provide a full specification for the layout-delimited literal form introduced by a forward reference, \li{~}, as well as other forms of forward-referenced blocks (for the forms \li{new} and \li{case(e)}, in particular).
%\item We detail the general mechanism for associating metadata with a type. A TSL is then implemented by associating a parser (of type \li{Parser}) with a type. The parser is responsible for rewriting tokenstreams (of type \li{Tokenstream}) into Wyvern ASTs (of type \li{Exp}). These types are defined in the standard library.
%\item This lower-level mechanism is general, but writing a hand-written parser and manipulating syntax trees manually is cognitively demanding. We observe that \emph{grammars} and \emph{quasiquotes} can both be seen as TSLs for parsers and ASTs respectively and discuss how to implement them as such.
\item We formalize the static semantics, including the literal parsing logic, of TSL Wyvern by combining a bidirectional type system (in the style of Lovas and Pfenning \cite{Lovas08abidirectional}) with an elaboration semantics (in a style similar to Harper and Stone \cite{Harper00atype-theoretic}). By distinguishing locations where an expression synthesizes a type from locations where an expression is being analyzed against a known type, we can precisely state where generic literals can and cannot appear and how parsing is delegated to a TSL.
\item A na\"ive rewriting strategy would be \emph{unhygienic} -- it could allow for the inadvertent capture of local variables. We show a novel mechanism that ensures hygiene by requiring that the generated AST is closed except for subtrees derived from portions of the user's tokenstream that are interpreted as nested Wyvern expressions. We also show how to explicitly refer to local values available in the parser definition (e.g. helper functions) in a safe way. We formalize this hygiene mechanism by bifurcating the context in our static semantics.
\item We provide several examples of TSLs throughout the paper, but to examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code  (collaborative work with Darya Kurilova).
\end{enumerate}

\subsection{Remaining Tasks and Timeline}
The paper we have submitted to ECOOP 2014 has received quite positive reviews, so we anticipate that it will be accepted (notifications are sent on Mar. 7th). If accepted, final revisions on the paper text are due on May 12th and so we will largely do these tasks after the OOPSLA deadline in late March. If not accepted, we will aim to resubmit to OOPSLA.
\begin{enumerate}
%\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints. This is not strictly necessary for preventing ambiguities, and it is quite convenient to not require this (as in Python, for example). We will add these simple rules to the grammar.
\item We must write down the full formal semantics (including rules that we omitted for concision in the paper draft) in a technical report, as well as provide more complete proofs of the metatheory. This might require slightly restricting the recursion in the rule for literals, so that induction is well-founded.
\item Our current static semantics does not support mutually recursive type declarations, but this is necessary for our prelude to typecheck, so we will add support for this. In the paper, the formalism is meant to be expository, so we will likely leave this additional complexity for the tech report.
\item We must further consider aspects of hygiene:
\begin{enumerate}
\item We do not yet have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}. It may be possible to prevent shadowing by making it impossible for variables to be introduced into interpolated Wyvern expressions directly (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
\item In macro systems like Scheme's, free variables in generated ASTs refer to their values within the macro definition, rather than where they are inserted. To support this, we have introduced the \li{valAST} operator. Using it is a bit verbose and awkward -- every free variable in a quasiquote must be inside \li{valAST}. We believe we can insert \li{valAST} automatically from within the definition of the TSL for \li{Exp} using a very simple extension to the interpolation mechanism that allows interpolation of single variables, rather than whole expressions, and plan to write up how this works.
\end{enumerate}
%\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
%\item The current implementation of Wyvern does not include all of these mechanisms as described. Although the implementation of Wyvern is not a project I am responsible for, I plan on working with the student who is doing this to implement as much of what we have described as possible.\todo{should I drop this from the proposal? I don't want to promise that everything will be implemented because that is largely contingent on Benjamin's effort, not mine.}
\end{enumerate}
%
%\subsection{TODO}
%- Figure out separate compilation (e.g. of CSS)
